The Forecasting Trap -- notes

### 2022-01-05 notebook/alternate-models.Rmd
alternate "model 1" example added

```{r}
# these dynamics are nonsense
# predator pop goes to Inf if prey is extinct
# prey = sp 1, predator= sp 2
R1 <- R2 <- 3
A1 <- A2 <- 2
B1 <- 0.5
B2 <- -0.3
Tmax <- 50
omega1 <- rnorm(Tmax, 0, 0.001)
omega2 <- rnorm(Tmax, 0, 0.001)

@@

df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)

```{r}
ggplot(df) +
 geom_path(aes(time, x1)) + 
 geom_path(aes(time, x2), col="red")
```

model_5sp <- function(u_prey = 0.45, u_pred = .66) {
    Tmax <- 50
    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
    xi1 <- rnorm(Tmax, 0, 0.01)
    xi2 <- rnorm(Tmax, 0, 0.01)
    xi3 <- rnorm(Tmax, 0, 0.01)
    xi4 <- rnorm(Tmax, 0, 0.01)
    xi5 <- rnorm(Tmax, 0, 0.01)

    x1[1] <- x2[1] <- x3[1] <- 1
    x4[1] <- x5[1] <- .5

df <- model_5sp()

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "population") %>%
    ggplot(aes(t, population, col= species)) + geom_path()
```

utility <- function(x) {
    delta <- 0.99
    w1 <- 0.25
    w2 <- 0.25
    w3 <- 1 - (w1 + w2)
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- model_5sp(u_prey, u_pred)
    R <- 
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * x5
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))

o <- optim(c(0.5,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

- Determine n alternate models over the same state space. Consider structurally different models, ideally with structures consistent with
  common forecast / inference models

>>>>

model_5sp <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfushed equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50) {
    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
    xi1 <- rnorm(Tmax, 0, sigma)
    xi2 <- rnorm(Tmax, 0, sigma)
    xi3 <- rnorm(Tmax, 0, sigma)
    xi4 <- rnorm(Tmax, 0, sigma)
    xi5 <- rnorm(Tmax, 0, sigma)

    x1[1] <- x1o
    x2[1] <- x2o
    x3[1] <- x3o
    x4[1] <- x4o
    x5[1] <- x5o

utility <- function(x) {
    delta <- 0.999
    w3 <- 0.5
    w2 <- 0.25
    w1 <- 1 - (w2 + w3)
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- model_5sp(u_prey, u_pred)
    R <- 
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))
-utility(c(0,0))

o <- optim(c(0.,0.), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

o <- optim(c(0.5,0.1), utility)
clip(o$par, 0, 1)
o$value

For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
As researchers, we only have access to data derived from observations from the model. 
Let us consider a sample simulation of such data:

```{r}
df <- model_5sp(0, .6, sigma = 0.01)

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(t, biomass, col= species)) + geom_path()
```

There are many other models, including process-based models, statistical models or even black-box machine learning models
which 5 dimensional state-space and two-dimensional action space.

```{r}
df <- model_5sp(0, 0., sigma = 0.01)

df %>%
    transmute(time = t, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```

A reasonable single-species model for Bass achieves the optimal solution.
The model is reasonable in the sense that it is a simple Ricker model 
that comparably reproduces the unfished dynamics of the Bass.
Nevertheless, it is actually quite poor at forecasting fished stock sizes.

model_1sp <- function(u = 0, A = 2, R = 1.6, Tmax=50, x1o = .8, sigma = 0.01) {

    omega1 <- rnorm(Tmax, 0, sigma)

    s1 <- x1 <- numeric(Tmax)
    x1[1] <- x1o

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - u)
        x1[t+1] <- s1[t] * exp(R - s1[t] * A + omega1[t])
    }

    df <- tibble(time = 1:Tmax, x1)
}

df <- model_1sp(0)
df %>% ggplot(aes(t, x1)) + geom_path()

```{r}
utility <- function(x) {
    delta <- 0.999
    u <- clip(x, 0, 1)
    df <- model_1sp(u)
    R <- u * df$x1 
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

o <- optimize(utility, c(0,1))
o$minimum
```

```{r}
df <- model_1sp(o$minimum)
df %>% ggplot(aes(t, x1)) + geom_path()
```

```{r}
R1 <- R2 <- 3
A1 <- A2 <- 2
B1 <- 0.5
B2 <- -0.3
Tmax <- 50
omega1 <- rnorm(Tmax, 0, 0.001)
omega2 <- rnorm(Tmax, 0, 0.001)

@@

df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)
```
